---
title: "Public Health and Economic Impact of Storm Types in the United States, 1950-2011"
author: "Cara Hubbell"
date: "February 9, 2016"
output: html_document
---

###Synopsis

Text here. No more than 10 sentences to describe analysis.

###Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

###Data Processing

In this report, we explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. The database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The data for this report come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. The data can be downloaded using the following link:

[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47 Mb]

The following code downloads and saves the data to the user's working directory, then loads the data into R. R packages are required here as well.

```{r, echo=TRUE}
require(dplyr)
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
  destfile="StormData.csv.bz2")
data<-read.csv("StormData.csv.bz2")
```

There is also some documentation of the database from the NOAA. The documentation shows how some of the variables are constructed and/or defined. The user may choose to read these documents; however, it is not necessary to understand the remainder of this report.

[National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  
[National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years are considered to be more complete.

Once the data set is loaded into R, we can see the structure of the data using:

```{r, echo=TRUE}
str(data)
```

The data set consists of 902297 observations. Each observation is one storm event. We use the terms "observation" and "event" interchangeably throughout this paper.

Since this is an analysis of types of storm events (coded in the variable ```EVTYPES```), the names of the values within ```EVTYPE``` are cleaned up slightly by making them entirely lower case. This brings the number of unique values for ```EVTYPE``` down nearly 10%. 

```{r, echo=TRUE}
data$EVTYPE<-tolower(data$EVTYPE)
```

*For a more rigorous analysis, the values for ```EVTYPE``` require further modifications, e.g. "winter mix", "wintery mix", "wintry mix", "winter weather mix", and "winter weather/mix" should be considered equivalent. For the purpose of this report, these modifications were not made.*

###Analysis 

First, we will investigate which types of storm events have had the greatest impact on humans. From ```str(data)``` and from the documentation files, we know that  the variables ```FATALITIES``` and ```INJURIES``` represent the total number of reported human deaths and injuries respectively for each storm event.

We now create a new variable called ```totalHarmed``` which is the sum of the number of fatalities and of injuries for each event.

```{r, echo=TRUE}
data$totalHarmed<-data$FATALITIES+data$INJURIES
```

Since we are interested in knowing the storm type (```EVTYPE```) that has caused the most harm to humans, we find the sum of ```totalHarmed``` within each ```EVTYPE``` and then reorder the data set with ```totalHarmed``` in descending order.

```{r, echo=TRUE}
data<-arrange(data, EVTYPE)
EVTYPES<-unique(data$EVTYPE)

x<-tapply(data$totalHarmed, data$EVTYPE, sum, na.rm=TRUE)
x<-data.frame(EVTYPE=EVTYPES, totalHarmed=as.numeric(x))
x<-arrange(x, desc(totalHarmed))
```

From ```str(x)``` we see that there are 898 different types of storm events. 

```{r, echo=TRUE}
str(x)
```

While it may be possible to represent all 898 types of storm events graphically, we will plot only the top six most harmful event types for humans. We are choosing six somewhat arbitrarily; this yields a nice, readable plot while also clearly showing what storm types are the most harmful for humans.

```{r, echo=TRUE}
xtop<-x[1:6,]

barplot(xtop$totalHarmed, 
        col=rainbow(6, s = 1, v = 1, start = .5, end = .75, alpha = 1),
        xlab="Event Type",
        ylab="Number of People Injured or Killed",
        main="Storm Events with the Greatest Human Impact"
        )
legend("topright",
       tolower(xtop$EVTYPE),
       cex=.6, 
       fill=rainbow(6, s = 1, v = 1, start = .5, end = .75, alpha = 1)
       )
```

Again, the top six most harmful storm event types for humans, in order of severity, are:
```{r, echo=TRUE}
xtop$EVTYPE
```
with the most injuries and deaths occurring as a result of:
```{r, echo=TRUE}
xtop$EVTYPE[1]
```

The next question we investigate is what storm event type has had the greatest economic impact. To analyze this, we consider ```PROPDMG``` and ```CROPDMG``` which provide the value of damage to property and agriculture respectively measured in US dollars. The variables ```PROPDMGEXP``` and ```CROPDMGEXP``` contain modifiers for the dollar values given in ```PROPDMG``` and ```CROPDMG```; "K" represents thousands of dollars, "M" represents millions of dollars, and "B" represents billions of dollars. Some values within the modifier variables are comments or other symbols. These values are ignored.

To analyze the total economic impact of each storm type, two new variables are created: ```totalProp``` which represents the total value of property damage measured in thousands of dollars and ```totalCrop``` which represents the total value of agricultural damage measured in thousands of dollars.

```{r, echo=TRUE}
data$totalProp<-0 
data$totalProp[which(data$PROPDMGEXP=="K")]<-data$PROPDMG[which(data$PROPDMGEXP=="K")]
data$totalProp[which(data$PROPDMGEXP=="M")]<-data$PROPDMG[which(data$PROPDMGEXP=="M")]*1000
data$totalProp[which(data$PROPDMGEXP=="B")]<-data$PROPDMG[which(data$PROPDMGEXP=="B")]*1000000


data$totalCrop<-0
data$totalCrop[which(data$CROPDMGEXP=="K")]<-data$CROPDMG[which(data$CROPDMGEXP=="K")]
data$totalCrop[which(data$CROPDMGEXP=="M")]<-data$CROPDMG[which(data$CROPDMGEXP=="M")]*1000
data$totalCrop[which(data$CROPDMGEXP=="B")]<-data$CROPDMG[which(data$CROPDMGEXP=="B")]*1000000
```

Next, the values of ```totalProp``` and ```totalCrop``` are combined to form the value of ```totalDamage```. This variable represents the total dollar value of all damage resulting from each observation measured in thousands of dollars.

```{r, echo=TRUE}
data$totalDamage<-data$totalProp+data$totalCrop
```

As previously stated, we are interested in knowing what storm event type (```EVTYPE```) has had the greatest economic impact. To determine this, the values of ```totalDamage``` are summed within each ```EVTYPE```. The event types are  then sorted in decreasing order (most severe first).

```{r, echo=TRUE}
y<-tapply(data$totalDamage, data$EVTYPE, sum, na.rm=TRUE)
y<-data.frame(EVTYPE=EVTYPES, totalDamage=as.numeric(y))
y<-arrange(y, desc(totalDamage))
```

Again, from ```str(y)``` we know that there are 898 different types of storm events. 

```{r, echo=TRUE}
str(y)
```

Just as when we considered the impact of storm types on humans, we will plot only the top six most damaging event types, even though it may be possible to graphically represent all event types. To reiterate, we are choosing six somewhat arbitrarily; this yields a nice, readable plot while also clearly showing what storm types cause the most economic damage.

```{r, echo=TRUE}
ytop<-y[1:6,]

barplot(ytop$totalDamage, 
        col=rainbow(6, s = 1, v = 1, start = .9, end = .1, alpha = 1),
        xlab="Event Type",
        ylab="Total Damage",
        main="Storm Events with the Greatest Economic Impact"
        )
legend("topright",
       tolower(ytop$EVTYPE),
       cex=.6, 
       fill=rainbow(6, s = 1, v = 1, start = .9, end = .1, alpha = 1)
       )
```

The plots shows that the top six most harmful storm event types for humans, in order of severity, are:
```{r, echo=TRUE}
ytop$EVTYPE
```
with the most severe economic impact occurring as a result of:
```{r, echo=TRUE}
ytop$EVTYPE[1]
```

###Results
From the analysis, it was shown that the storm types with the most significant negative impact on humans in the United states were:
```{r, echo=TRUE}
xtop
```
and that the storm types causing the greatest economic damage were:
```{r, echo=TRUE}
ytop
```
We can see that there is some overlap between these two subsets of storm event types. Tornados and floods are both among the top four most deadly storm events for human lives and for property/agricultural goods. These results also show that tornados have killed and injured more than ten times as many people as the next most significant storm event type and floods have caused more than twice as much monetary damage as the next most significant type.

###Conclusions and Limitations

This paper does not seek to make any suggestions as far as preventing loss of lives or monetary damage due to various types of storm events. We merely show which weather incidents have inflicted the most devastation in the United States between 1950 and 2011.

While the results appear reasonable, namely that tornados and floods are among the most destructive, the actual counts for humans injured/killed and the economic dollar value of damage to property/crops should not be taken as precise measurements. This is because of the precision with which data was recorded within the NOAA database. ```EVTYPE``` has 985 unique values in the raw data set. Modifying these values to be entirely lower case dropped the number of unique values to 898, which is a decrease of nearly 10%. However, several of the values clearly represent the same type of event (e.g. "urban and small stream", "urban and small stream flood", "urban and small stream floodin", "urban/small stream  flood", "urban/sml stream fld", etc.). Neither the data cleaning nor the analysis itself took this issue into account. Fixing this issue and cleaning up ```EVTYPE``` values could have a significant impact on the data

